---
title: "Your BBOB Fitness Function May Be Burning More Energy Than You Think"
subtitle: "Key findings from *Measuring Energy Consumption of BBOB Fitness Functions* (EvoStar 2025)"
author: "Juan J. Merelo-Guervós, Gustavo Romero López, Mario García-Valdez"
date: "2025-04-25"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    fig_caption: true
    df_print: paged
bibliography: ../paper/ours.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.align = "center", out.width = "100%")
# Install optional packages if not present
for (.pkg in c("ggrepel", "scales", "tidyr")) {
  if (!requireNamespace(.pkg, quietly = TRUE)) {
    install.packages(.pkg, repos = "https://cloud.r-project.org", quiet = TRUE)
  }
}
library(ggplot2)
library(ggthemes)
library(dplyr)
library(knitr)
library(scales)
```

---

## The Hidden Cost of Evaluating Solutions

Every time an evolutionary algorithm evaluates a candidate solution, it runs a
**fitness function** — usually dozens of thousands of times per experiment, and
thousands of experiments over a research career. We rarely stop to ask: *how
much energy does that actually cost?*

In a world where [software is responsible for a growing share of global
electricity consumption](https://doi.org/10.1007/978-3-031-90065-5_15),
understanding and minimising the energy footprint of our algorithms is no longer
optional. It is part of good engineering — and good science.

In our paper published at
[EvoStar 2025](https://link.springer.com/chapter/10.1007/978-3-031-90065-5_15),
we asked a deceptively simple question:

> **Among the standard BBOB floating-point fitness functions, how much energy
> does each one consume — and do the representation choices we make (precision,
> data structure) actually matter?**

The answers were surprising.

---

## How We Measured It

We used [**pinpoint**](https://github.com/osmhpi/pinpoint), a multiplatform
tool that hooks into the CPU's built-in power sensors (Intel/AMD RAPL) and
reports the total energy consumed by the *package* — CPU cores plus memory —
during a process run. This is a hardware-level measurement, not a simulation.

The measurement procedure:

1. **Baseline**: generate 40 000 random floating-point chromosomes and record
   the energy (`PKG`, in Joules) consumed. This represents the unavoidable
   overhead of just *having* a population.
2. **Function evaluation**: apply each BBOB fitness function to the same 40 000
   chromosomes and record energy again.
3. **Delta**: subtract the baseline mean to isolate the energy cost of the
   function itself (`ΔE = PKG_function − mean(PKG_baseline)`).

We repeated every configuration **30 times** on a dedicated AMD Ryzen 9 3950X
machine (Ubuntu 20.04, g++ 10.5, `-O3 -march=native -flto`), varying:

- **Precision**: `float` (32-bit) vs `double` (64-bit)
- **Chromosome size**: 128, 256 and 512 dimensions
- **Data structure**: variable-size `std::vector` vs fixed-size `std::array`

Ten BBOB functions were chosen to span all structural categories: `sphere`,
`rastrigin`, `rosenbrock`, `discus`, `bent_cigar`, `different_powers`,
`sharp_ridge`, `schaffers`, `schwefel`, and `katsuura`.

---

## Finding 1: The Energy Landscape Spans Orders of Magnitude

```{r load-data}
# Load and prepare baseline data
base.variable.data <- read.csv("../data/evostar25-generation-5-Nov-07-30-15.csv")
base.variable.data$size <- as.factor(base.variable.data$size)
base.variable.data$type <- as.factor(base.variable.data$type)
base.variable.data %>%
  group_by(size, type) %>%
  summarise(mean.PKG = mean(PKG), .groups = "drop") -> summary.base.variable.data

# Load and prepare function evaluation data
functions.data <- read.csv("../data/variable-evostar25-bbob-10-Nov-19-10-32.csv")
functions.data <- functions.data[functions.data$work != "none", ]
functions.data$delta.PKG <- 0

for (sz in c(128, 256, 512)) {
  for (tp in c(" f", " d")) {
    mask     <- functions.data$size == sz & functions.data$type == tp
    baseline <- summary.base.variable.data[
      summary.base.variable.data$size == sz &
        summary.base.variable.data$type == tp, ]$mean.PKG
    functions.data[mask, ]$delta.PKG <-
      functions.data[mask, ]$PKG - rep(baseline, sum(mask))
  }
}
functions.data$delta.PKG <- pmax(functions.data$delta.PKG, 0)

# Clean up type labels for plotting
functions.data$precision <- ifelse(trimws(functions.data$type) == "f", "float (32-bit)", "double (64-bit)")
```

```{r fig-all-functions, fig.cap="**Figure 1.** Median energy consumption (ΔE, Joules, log scale) of each BBOB fitness function for 40 000 evaluations, by chromosome size and floating-point precision. Functions are ranked by their median energy at size 512 with double precision. The y-axis is on a base-10 log scale — each tick is a 10× increase."}
# Compute medians for ranking
rank_order <- functions.data %>%
  filter(size == 512, trimws(type) == "d") %>%
  group_by(work) %>%
  summarise(med = median(delta.PKG), .groups = "drop") %>%
  arrange(med) %>%
  pull(work)

functions.data$work <- factor(functions.data$work, levels = rank_order)
functions.data$size_label <- paste0("Size = ", functions.data$size)
functions.data$size_label <- factor(functions.data$size_label,
                                     levels = c("Size = 128", "Size = 256", "Size = 512"))

ggplot(functions.data, aes(x = work, y = delta.PKG, colour = precision)) +
  geom_boxplot(outlier.size = 0.8, outlier.alpha = 0.5) +
  facet_wrap(~size_label, ncol = 1) +
  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x))) +
  scale_colour_brewer(palette = "Set1", name = "Precision") +
  labs(
    x = "BBOB function",
    y = "ΔE (Joules, log scale)"
  ) +
  theme_bw(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 40, hjust = 1),
    strip.background = element_rect(fill = "grey90"),
    legend.position = "top"
  )
```

The first thing that jumps out from Figure 1 is the **extraordinary spread in
energy consumption**. Using a logarithmic scale is not a stylistic choice — it
is a necessity: `katsuura` consumes roughly **one order of magnitude more
energy** than most other functions at the same chromosome size and precision.

To put that in concrete terms:

```{r summary-table}
functions.data %>%
  group_by(work, precision, size) %>%
  summarise(
    median_J  = round(median(delta.PKG), 4),
    q75_J     = round(quantile(delta.PKG, 0.75), 4),
    .groups   = "drop"
  ) %>%
  filter(size == 512) %>%
  select(-size) %>%
  arrange(precision, median_J) %>%
  rename(Function = work, Precision = precision,
         `Median ΔE (J)` = median_J, `75th pctile ΔE (J)` = q75_J) %>%
  kable(caption = "Table 1. Energy cost per batch of 40 000 evaluations at chromosome size 512.")
```

Some functions (`bent_cigar`, `discus`, `rosenbrock` at small sizes) show
**zero measurable delta**. This is *not* a mistake — the compiler's aggressive
optimiser folds these lightweight functions into the chromosome generation loop,
making them energetically inseparable from the baseline. That is good news for
users of those functions, but it also means RAPL-based measurement hits a
precision floor of roughly 5% of total PKG consumption.

---

## Finding 2: Double Precision Is Not the Enemy

The conventional wisdom says *"use floats to save energy"* — halving the
byte width should halve the memory traffic and therefore the energy, right?

**Not quite.**

```{r fig-float-vs-double, fig.cap="**Figure 2.** Head-to-head comparison of float vs double at chromosome size 512 (variable-size vector). A position above the diagonal means double costs more; below means float costs more. Most functions cluster near the diagonal or below it."}
comparison <- functions.data %>%
  filter(size == 512) %>%
  group_by(work, precision) %>%
  summarise(median_dE = median(delta.PKG), .groups = "drop") %>%
  tidyr::pivot_wider(names_from = precision, values_from = median_dE) %>%
  rename(float32 = `float (32-bit)`, double64 = `double (64-bit)`)

ggplot(comparison, aes(x = double64, y = float32, label = work)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
  geom_point(size = 3, colour = "#E41A1C") +
  ggrepel::geom_text_repel(size = 3.5, max.overlaps = 20) +
  scale_x_log10(labels = scales::trans_format("log10", scales::math_format(10^.x))) +
  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x))) +
  labs(
    x = "double (64-bit) median ΔE (J)",
    y = "float (32-bit) median ΔE (J)"
  ) +
  theme_bw(base_size = 12)
```

In Figure 2, points on the **diagonal** indicate identical energy cost for both
precisions. Points **below** the diagonal mean `float` is *more expensive* than
`double` — the opposite of what intuition suggests.

Why? Modern compilers, when instructed with `-O3 -march=native`, generate
vectorised 64-bit SIMD code that processes doubles very efficiently. Reducing to
32-bit floats does not automatically unlock a cheaper code path; in many cases
the same instructions are used, but the shorter float values have to be
zero-extended or handled with narrower SIMD widths, negating the theoretical
gain. The **only function where float consistently wins is `katsuura`**, where
the sheer amount of trigonometric and power operations means fewer bits do
translate to measurably less computation.

> **Practical takeaway**: there is no blanket energy argument for switching your
> EA from `double` to `float`. The theoretical gain rarely materialises in
> practice, and you lose several digits of precision.

---

## Finding 3: Where Arrays Really Shine — and Where They Don't

We also tested a fixed-size `std::array<T, 128>` against the usual
variable-size `std::vector<T>`.

```{r load-fixed-data}
base.fixed.data <- read.csv("../data/fixed-evostar25-generation-5-Nov-07-37-13.csv")
base.fixed.data$size  <- as.factor(base.fixed.data$size)
base.fixed.data$type  <- as.factor(base.fixed.data$type)
base.fixed.data %>%
  group_by(size, type) %>%
  summarise(mean.PKG = mean(PKG), .groups = "drop") -> summary.base.fixed.data

fixed.functions.data <- read.csv("../data/evostar25-bbob-fixed-12-Nov-08-20-03.csv")
fixed.functions.data$delta.PKG <- 0
for (tp in c(" f", " d")) {
  mask <- fixed.functions.data$size == 128 & fixed.functions.data$type == tp
  baseline <- summary.base.fixed.data[
    summary.base.fixed.data$size == 128 &
      summary.base.fixed.data$type == tp, ]$mean.PKG
  fixed.functions.data[mask, ]$delta.PKG <-
    fixed.functions.data[mask, ]$PKG - rep(baseline, sum(mask))
}
fixed.functions.data$delta.PKG <- pmax(fixed.functions.data$delta.PKG, 0)
fixed.functions.data$data.structure <- "array (fixed)"
functions.data$data.structure       <- "vector (variable)"
fixed.functions.data$precision <- ifelse(trimws(fixed.functions.data$type) == "f",
                                          "float (32-bit)", "double (64-bit)")

fixed.functions.data$size_label <- paste0("Size = ", fixed.functions.data$size)
fixed.functions.data$size_label <- factor(fixed.functions.data$size_label,
                                     levels = c("Size = 128", "Size = 256", "Size = 512"))

functions.128 <- rbind(
  fixed.functions.data[fixed.functions.data$size == 128, ],
  functions.data[functions.data$size == 128, ]
)
functions.128$work <- factor(functions.128$work, levels = rank_order)
```

```{r fig-generation-baseline, fig.cap="**Figure 3.** Energy consumed by chromosome *generation* alone (the unavoidable baseline), comparing fixed-size `array` vs variable-size `vector` at size 128. Both float and double are shown. The array baseline is roughly 8× cheaper.", fig.height=4}
base.variable.data.128 <- read.csv("../data/evostar25-generation-5-Nov-07-30-15.csv")
base.variable.data.128 <- base.variable.data.128[base.variable.data.128$size == 128, ]
base.variable.data.128$data.structure <- "vector (variable)"
base.fixed.data.plot <- read.csv("../data/fixed-evostar25-generation-5-Nov-07-37-13.csv")
base.fixed.data.plot$data.structure  <- "array (fixed)"
base.gen <- rbind(base.variable.data.128, base.fixed.data.plot)
base.gen$precision <- ifelse(trimws(base.gen$type) == "f", "float", "double")

ggplot(base.gen, aes(x = data.structure, y = PKG, colour = precision, fill = precision)) +
  geom_violin(alpha = 0.3, trim = FALSE) +
  geom_point(position = position_jitter(width = 0.1), alpha = 0.5, size = 1.5) +
  scale_colour_brewer(palette = "Set1", name = "Precision") +
  scale_fill_brewer(palette = "Set1", name = "Precision") +
  labs(
    x = "Data structure",
    y = "PKG energy (Joules)"
  ) +
  theme_bw(base_size = 12) +
  theme(legend.position = "top")
```

The **array baseline is approximately 8× cheaper** than the vector baseline
(Figure 3). This is significant: every evolutionary algorithm spends time at
initialisation creating chromosomes, and if those chromosomes are fixed-size,
the savings at startup are real and large.

However, for *function evaluation* the picture is murkier:

```{r fig-array-vs-vector, fig.cap="**Figure 4.** ΔE for function evaluation at size 128, comparing array and vector data structures. Double precision shown; the results for float are analogous. For most functions the two data structures are equivalent — the `katsuura` exception aside.", fig.height=4}
ggplot(functions.128[functions.128$precision == "double (64-bit)", ],
       aes(x = work, y = delta.PKG, colour = data.structure)) +
  geom_boxplot(outlier.size = 0.8, outlier.alpha = 0.4, position = "dodge") +
  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x))) +
  scale_colour_manual(values = c("array (fixed)" = "#4DAF4A", "vector (variable)" = "#984EA3"),
                      name = "Data structure") +
  labs(
    x = "BBOB function",
    y = "ΔE (Joules, log scale)"
  ) +
  theme_bw(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 40, hjust = 1),
    legend.position = "top"
  )
```

For most functions there is **no meaningful difference** in evaluation energy
between the two structures. The compiler emits essentially the same inner loop
regardless of whether the data pointer comes from a `vector` or an `array`. The
conclusion is clear:

> **Use `double + vector` as your default.** It is the combination that requires
> no compile-time size specialisation and consistently delivers the lowest or
> tied-lowest evaluation energy across functions. Switch to `array` only if
> your algorithm initialises large populations frequently — the 8× saving
> at chromosome-generation time can add up over an evolutionary run.

---

## Going Further: Power Profile Across Functions

The paper's *future work* section mentions a need to look beyond raw Joules and
consider **average power** (watts = J/s), which characterises how
*intensively* a function taxes the CPU rather than just how long it runs. We
can do this directly with the existing data.

```{r fig-watts-profile, fig.cap="**Figure 5.** Average power draw (watts = PKG / seconds) for each BBOB function at size 512, double precision. This reveals which functions are *computationally intense* regardless of how long they take. Functions are ordered by median power.", fig.height=4.5}
functions.data.watts <- read.csv("../data/variable-evostar25-bbob-10-Nov-19-10-32.csv")
functions.data.watts <- functions.data.watts[functions.data.watts$work != "none", ]
functions.data.watts$watts <- functions.data.watts$PKG / functions.data.watts$seconds
functions.data.watts$precision <- ifelse(trimws(functions.data.watts$type) == "f",
                                          "float (32-bit)", "double (64-bit)")

watts_order <- functions.data.watts %>%
  filter(size == 512, trimws(type) == "d") %>%
  group_by(work) %>%
  summarise(med_w = median(watts), .groups = "drop") %>%
  arrange(med_w) %>%
  pull(work)

functions.data.watts$work <- factor(functions.data.watts$work, levels = watts_order)

ggplot(functions.data.watts[functions.data.watts$size == 512, ],
       aes(x = work, y = watts, colour = precision)) +
  geom_boxplot(outlier.size = 0.8, outlier.alpha = 0.5) +
  scale_colour_brewer(palette = "Set1", name = "Precision") +
  labs(
    x = "BBOB function",
    y = "Average power (watts)"
  ) +
  theme_bw(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 40, hjust = 1),
    legend.position = "top"
  )
```

Figure 5 reveals something the raw-energy view obscures: while `katsuura` still
tops the ranking, the **relative ordering of the other functions changes**. A
function that consumes modest total Joules but finishes quickly may draw more
*instantaneous* watts than one that consumes more Joules over a longer run. This
matters for thermal management and for cloud environments billed by peak power.

---

## Energy Scaling with Chromosome Size

One more analysis the existing data allows us to perform is to see **how energy
scales** as chromosome size grows from 128 to 256 to 512 dimensions. Theory
predicts linear scaling for simple element-wise functions, but loop-heavy or
transcendental functions might scale super-linearly.

```{r fig-scaling, fig.cap="**Figure 6.** Median ΔE as a function of chromosome size, for each BBOB function (double precision, variable vector). Lines connect medians across sizes. Functions with near-zero measurements at size 128 are excluded from the log scale.", fig.height=5}
scaling <- functions.data %>%
  filter(precision == "double (64-bit)", delta.PKG > 0) %>%
  group_by(work, size) %>%
  summarise(med_dE = median(delta.PKG), .groups = "drop")

ggplot(scaling, aes(x = size, y = med_dE, colour = work, group = work)) +
  geom_line(size = 0.9) +
  geom_point(size = 2.5) +
  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x))) +
  scale_colour_brewer(palette = "Set3", name = "Function") +
  labs(
    x = "Chromosome size (dimensions)",
    y = "Median ΔE (Joules, log scale)"
  ) +
  theme_bw(base_size = 12) +
  theme(legend.position = "right")
```

The scaling plot shows that **all functions grow roughly linearly** in energy
with chromosome size on this log-log scale — that is, a doubling of size leads
to roughly a doubling of energy. There is no evidence of any function becoming
disproportionately expensive at larger sizes; the rank order is stable. This is
reassuring: measurements taken at size 128 give a reliable relative picture even
for larger problems.

---

## Summary: What Should You Actually Do?

| Decision | Recommendation | Confidence |
|---|---|---|
| `float` vs `double` | Use `double` | High — float rarely saves energy and costs precision |
| `vector` vs `array` | Use `vector` for evaluation; `array` for init-heavy workloads | High |
| Which functions to benchmark? | Prefer `katsuura`, `rastrigin`, `schaffers` — they are measurable | High |
| Chromosome size | Measurements at 128 dimensions generalise to larger sizes | Medium |
| Power vs energy | Monitor both; they do not always rank functions the same way | Medium |

---

## Open Questions and Future Work

The paper identified several open questions that remain fertile ground for
further investigation:

1. **Compiler-fused functions**: `bent_cigar`, `discus`, `rosenbrock` (at small
   sizes) produce zero delta energy because the compiler merges them with the
   generation loop. Techniques like `volatile` fencing or separate compilation
   units could make these measurable.

2. **Reduced precision beyond `float`**: The transprecision computing literature
   suggests 8-bit and 16-bit floating-point formats can cut energy by up to 30%
   in some workloads. Could BBOB functions tolerate such reduced precision? The
   answer likely varies by function — `katsuura` with its iterated products
   might be sensitive, while `sphere` could be fine.

3. **Other BBOB functions**: 14 of the 24 BBOB functions remain unmeasured. Do
   any of them approach `katsuura`'s extreme energy footprint?

4. **Other architectures**: All results here are for an AMD Ryzen on Linux.
   Intel, ARM and RISC-V CPUs vectorise differently, and the
   float-vs-double story may be completely different on an M-series Apple chip.

5. **Algorithm-level impact**: The paper measures function calls in isolation.
   A full evolutionary algorithm adds selection, crossover, mutation, and
   population management. How does function energy cost translate to total run
   energy?

---

## References

This post is based on:

> Merelo-Guervós, J. J., Romero López, G., & García-Valdez, M. (2025).
> *Measuring Energy Consumption of BBOB Fitness Functions.*
> In *Applications of Evolutionary Computation* (EvoStar 2025),
> Lecture Notes in Computer Science, vol. 15613, pp. 240–254. Springer.
> <https://doi.org/10.1007/978-3-031-90065-5_15>

Source code and data: <https://github.com/JJ/energy-bbob> (GPL licence).
